---
title       : Session4
subtitle    : 
author      : Mandy
job         : 
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : default      # default tomorrow 
widgets     : [mathjax,bootstrap]            # {mathjax, quiz, bootstrap}
mode        : standalone # {standalone, draft, selfcontained}
---

<style>
strong {
  font-weight: bold;
  color: red;
  font-size: 115%
}
</style>


--- .segue bg:grey

# News


--- 

## News


<iframe src="http://user2016.org/" width=100% height=100% allowtransparency="true"> </iframe>


--- &twocol w1:40% w2:60%

## News


*** =left

<iframe src="http://a2.typepad.com/6a0105360ba1c6970c01bb0885deea970d-pi" width=50% height=90% allowtransparency="true"> </iframe>

*** =right
Ross says that that observing the global labour of love has tempered his cynicism. 

<q2>R changed my opinion of humanity to some extent, to see how people are really willing to **freely give** of themselves and produce something larger than themselves **without** any thought of **personal glory**. There’s a **lot of work** with **no recognition.**</q2>


--- .segue bg:grey

# Recap


---

## Recap

You should know now:
>  - that statistics is all about simplifying 
>  - we try to summarize and describe data through parameters
     - location parameters (which? resp R commands?)
     - scale parameters/parameters of spread (which? resp R commands?)
 

--- 

## Recap

>- *mean()*, *median()*, *quantile()*
>- *sd()*, *range()*, *IQR()*


---


## Recap

We have seen how
>  - a difference of means (of two groups) 
>  - a spread parameter (standard deviation) and
>  - the sample sizea (as a parameter of uncertainty)

>  are combined to measure the difference of the location of two groups
>  this measure is compared to a null distribution through this comparison transformed in to a probability (p-value)


--- 

## test statistic

$$t = \frac{\bar{X}_{male} - \bar{X}_{female}}{s_{overall}\sqrt{\frac{1}{n_1} + \frac{1}{n_2}}}$$


--- 

## Two sided alternative

```{r,echo=FALSE,fig.height=8,fig.width = 12,fig.align='center'}
curve(dt(x,df=1000),from = -4,to = 4)
text(-4,0.35,"alternative: two sided",pos=4)
abline(h=0)
x <- seq(-4,qt(0.025,df=1000),by=0.01)
y <- dnorm(x)
x <- c(x,qt(0.025,df=1000))
y <- c(y,dt(-4,df=1000))
polygon(x,y,col="red")

x <- seq(qt(0.975,df=1000),4,by=0.01)
y <- dt(x,df=1000)
x <- c(x,qt(0.975,df=1000))
y <- c(y,dt(4,df=1000))
polygon(x,y,col="red")

text(-2.75,0.02,expression(paste("reject ",H[0])),pos=4)
text(1.9,0.02,expression(paste("reject ",H[0])),pos=4)
```


---

## One sided alternative - less

```{r, echo=FALSE,fig.height=8,fig.width=12,fig.align='center'}
curve(dt(x,df=1000),-4,4)
abline(h=0)
x <- seq(-4,qt(0.05,df=1000),by=0.01)
y <- dt(x,df=1000)
x <- c(x,qt(0.05,df=1000))
y <- c(y,dt(-4,df=1000))
polygon(x,y,col="red")

text(-4,0.35,"alternative: less",pos=4)
text(-2.45,0.02,expression(paste("reject ",H[0])),pos=4)

```

--- 

## One sided alternative - greater

```{r, echo=FALSE,fig.height=8,fig.width=12,fig.align='center'}
curve(dt(x,df=1000),-4,4)
abline(h=0)

x <- seq(qt(0.95,df=1000),4,by=0.01)
y <- dt(x,df=1000)
x <- c(x,qt(0.95,df=1000))
y <- c(y,dt(4,df=1000))
polygon(x,y,col="red")

text(-4,0.35,"alternative: greater",pos=4)
text(1.6,0.02,expression(paste("reject ",H[0])),pos=4)
```


--- .nobackground .quote 

<q>The p-value is the **probability** of the __sample estimate__ (of the respective estimator - here our *t*) __under the null__. We do not know anything of probabilities of the null or of estimates under the alternative!</q>


--- &vcenter &hcenter

## Decisions may be right - or wrong

|           | $H_0$ is true | $H_0$ is false|
|-----------|---------------|---------------|
|$H_0$ is not rejected| Correct decision|Type II error|
|$H_0$ is rejected| Type I error |  Correct decision|


--- .segue bg:grey

# T-tests in R

---

## T-tests in R

> - the one sample t-test
> - the two sample t-test assuming equal variances
> - the two sample t-test without the former assumption
> - the paired t-test (in fact: this is a one sample t-test against 0)
> - and there are others
> But: there is only one command in R: `t.test()`


---

## One Sample t-test

```{r}
set.seed(1)
x <- rnorm(12) ## create random numbers
t.test(x,mu=0) ## population mean 0
```


---

## One Sample t-test

```{r}
t.test(x,mu=1) ## population mean 1
```


--- 

## Two Sample t-test

>  - we have given two numeric vectors
>  - we do not assume equal variance for the underlying distributions


```{r}
set.seed(1)
x <- rnorm(12) ## create random numbers
y <- rnorm(12) ## create random numbers
head(data.frame(x,y))
```


--- 

## Two Sample t-test

```{r}
t.test(x,y)
```


---

## Two Sample t-test

>  - we have one numeric vector and one vector containing the group information
>  - we do not assume equal variance for the underlying distributions


```{r}
g <- sample(c("A","B"),12,replace = T) ## create random group vector
head(data.frame(x,g))
```


---

## Two Sample t-test

```{r}
t.test(x~g)
```


---

## Two Sample t-test

  - we assume equal variance for the underlying distributions

```{r}
t.test(x,y,var.equal = T)
```


---

## When should you use a t-test

>  - comparison of mean values against a null-value or against each other
>  - the t-test, especially the Welch test is appropriate whenever the underlying distributions are normal
>  - it is also recommended for group sizes $\geq 30$ (robust against deviation from normality)


---

## Exercises

Use the ALLBUS data set:
- do a t-test of income (V417): male against female (V81)!
- and compare the bmi (V279) in smokers and non-smokers (V272) and between people with high and normal blood pressure (V242)
- how would you interpret the results?
- visualize!


--- .segue bg:grey

# Rolling the dice

---

## Rolling the dice

Suppose you are rolling a fair die 600 times: what is the number of 6s you are expecting?

And how many 6s do we need to reject the NULL (a fair die)??

```{r}
qbinom(p = c(0.025,0.975),size = 600, prob = 1/6)
```


---

## Rolling the dice

So now let R rolling the dice!

```{r}
sample(1:6,600,replace = T)
```


---

## Rolling the dice

Count the 6s!

```{r}
sample(1:6,600,replace = T)==6
```


---

## Rolling the dice

Count the 6s!

```{r}
sum(sample(1:6,600,replace = T)==6)
```


---

## Exercise

Now use the following code to replicate the experiment (rolling one fair dice 600 times) 1000 times. The number of 6s are stored in the vector `x`.

How many statistically significant results do you expect?

How many statistically significant results did you get? (you can use `table()` in combination with a logical function)

Visualize the result using ggplot2 and `geom_histogram()` (look at the help of `geom_histogram` )! 

```{r}
x <- replicate(1000, sum(sample(1:6,600,replace = T)==6))
df <- data.frame(repid = 1:1000, n.6s = x)
head(df,2)
```


